Debugging by \textit{lastChange}

### 2010-10-26 papers/pldi2011/sigplanconf-template.tex
rough draft of evaluation section

>>>>

We have presented the \texttt{lastChange} algorithm and describe our
prototype implementation. Now we want to convince you that the
algorithm has practical value. Ultimately, practical value can only be
proven by developers in the field using the technique. This paper is
one step to convince implementers to put \texttt{lastChange} in the
field.  Thus our goal here is a persausive argument.

The key ingredients in our argument are: 
\begin{enumerate}
   \item developers need an operation like \texttt{lastChange}, 
   \item developers can learn to use \texttt{lastChange}, 
   \item practical implementations are feasible with modest
invested development time,
   \item in most cases \texttt{lastChange} will be much faster than current alternatives,
   \item the worst cases are not more common or more painful than
alternatives.
\end{enumerate}

\subsection{developers need an operation like \protect\texttt{lastChange} }

We have shown how \texttt{lastChange} identifies the prior point in
program execution where a program state value changes. Is it something
the developer needs to do?  Since ultimately programs are just
transformation of state values, debugging is ultimately backtracking
to find defects in program state change.  When a developer halts a
program on a breakpoint they have three kinds of information: a model
of the program in their head, a call stack showing what part of the
program is halted, and the state at that point (a combination of the
debugger's view of the program internal state and the output of the
program up to this point of execution). If the model does not match
the call stack or if no value in the viewable program state aligns
with the developers model for this point in execution, the developer
will seek another view by changing the breakpoint or the input
data. If a value appears incorrect, they may have a flash of insight
and know the defect. Otherwise they need to figure out what operation
causes the incorrect value. Thus \texttt{lastChange} addresses a key
part of the debugging process.

\subsection{developers can learn to use \protect\texttt{lastChange}}

If \texttt{lastChange} could be useful, can developers figure out how
to use it? We hope that our prototype demonstrates that
\texttt{lastChange} can be easily activated by operations on the
graphical representation of erroneous data in a debugger. Intepreting
the results should be only slightly more difficult. We designed the
user interface for the results from \texttt{lastChange} to resemble
the results from a breakpoint, with the source code of the change
point highlighted and the state of the program presented the way state
is presented from a breakpoint. Based on this user interface, we
believe developers can start to use \texttt{lastChange} with minimal
training. Of course developers may initially confuse the result of
\texttt{lastChange} with being halted in a debugger at the point of
last change. The value of the result will motivate them to clarify the
issue.

\subsection{practical implementations are feasible}

We described our prototype implementation in Sec. 4. A fully usable
implementation would require access to the object id at the point of
object creation and \texttt{setPropertyChangeHook()}. Many
object-oriented runtimes provide object identifiers\cite{} and some
provide access to object creation\cite{}. In the particular case of
the Firefox Web browser we used for our prototype, the primary barrier
to practical implementation would likely be integration with the
increasingly sophisticated just in time compilers\cite{}.

\subsection{in most cases \protect\texttt{lastChange} will be much faster than current alternatives}

When developers try \texttt{lastChange}, will they get results fast
enough to benefit? While we only used our prototype on toy programs,
\textit{lastChange} seemed as fast as re-execution.  Recall that we
insert additional code through debugger callbacks, then re-execute the
program. The additional code we insert is proportional (in our
JavaScript algorithm) to 1) the number of places a property or
variable with a given name is changed, 2) the number of places objects
are created. (Something about the likely overhead in Firefox).

For comparison we should use the practical alternative: developers
setting breakpoints. For the vast majority of programs, a developer
will take much more time to set one breakpoint than
\texttt{lastChange} would add. But typically the developer may not
guess the point of last change. They must then ponder another
breakpoint, and re-execution.

We could also compare to solutions based on logging or tracing. Manual
logging has very high overhead: the developer must add code, debug
that added code, then analyze the log. (To be fair, the log can become
a permenant debugging aid.) Automatic logging as we discussion in
\secref{related} causes about one or two orders of magnitude slow down.

\subsection{the worst cases are not more common or more painful than alternatives}

Finally we consider the worst cases: what about code that changes
objects in long running loops? Every time through the loop we incur
the call back overhead; if the loop itself has relatively little code
the overhead could be very large; if the loop computation is a
significant fraction of the full program, the slowdown would be
enormous.

Of course breakpoints are not feasible in these cases and logging
becomes unyieldy. Since automatic logging solutions are highly tuned,
their overhead in this case will likely be much less than
\texttt{lastChange}. On the other hand, \texttt{lastChange} integrates
with an interactive debugger and detecting that we are in a high
overhead loop is simply a matter of checking our internal counter. In
such unusual cases we may simply offer the developer the option of
studying the loop code then omitting it from \texttt{lastChange}
operations. As a practical result, this is no different from similar
issues developers face with any debugger today: occasionally a
debugger causes too much overhead to be useful for debugging. On
balance we think \texttt{lastChange} will rarely encounter these kinds
of issues.
